A Guide to Pre-Processing High-Throughput Animal Tracking Data

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
listings

An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 3.a).
Point outliers are characterised by artificially high speeds between the outlier and the positions before and after \citep[called incoming and outgoing speed, respectively][]{bjorneraas2010}, lending a `spiky' appearance to the track.
Removing spikes is simple: remove positions with extreme incoming and outgoing speeds.
Users must first define plausible upper limits for speed and turning angle for the study species \citep{calenge2009, seidel2018}.
Here, it is important to remember that speed estimates are scale-dependent; high-throughput tracking typically overestimates the speed between positions where the animal is stationary or moving slowly due to small-scale location errors \citep{ranacher2016, noonan2019}. 
Even after data with large location errors have been removed by filters, it is advisable to begin with a liberal (high) speed threshold that excludes only the most unlikely of speeds.
Estimates of maximum speed may not always be readily obtained for all species, and an alternative is to use a data-driven threshold such as the 95\textsuperscript{th} percentile of speeds from the track.
Once a speed threshold $S$ has been chosen, positions with incoming \textit{and} outgoing speeds $\geq S$ may be identified as spikes and removed.
Some species can realistically achieve speeds $\geq S$ in fast transit segments when assisted by environmental factors, such as birds with tailwinds, and a simple filter on incoming and outgoing speeds would exclude this valid data.
To avoid removing real, fast transit segments while still excluding spikes we suggest combining the speed filter with a filter on the turning angles of each position \citep[see][]{calenge2009}.
This combined filter assumes that positions in high-throughput tracking with both high speeds and large turning angles are likely to be due to location errors, since most species are unable to turn sharply at high speed.
Users can then retain only those positions whose incoming and outgoing speeds are both $< S$ which or satisfy the condition $\theta < A$, where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
The removal of spikes is implemented using the \texttt{atl\_filter\_covariates} function (Listing 4).
Many other track metrics may be used to identify implausible movement and on which data may be filtered \citep{seidel2018}.

@@

Sometimes entire susbets of the track may be affected by the same large-scale location error.
For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 3.a, b).
These cannot be corrected by targeted removal of individual positions, as in Bjorneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
Since filtering individual positions will not suffice, algorithms to correct such errors must take a track-level view, and target the displaced sequence overall.
Track-subset algorithms are likely to be system-specific, and may be challenging to conceptualise or implement.
In the case of prolonged spikes, one relatively simple solution is identifying the bounds and removing positions between them.
We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
Users are strongly encouraged to visualise their data before and after applying this method, and we caution against relying on this method if data are heavily distorted by errors affecting entire track-subsets.

@@

Smoothing does not change the number of observations, but does decouple the coordinates from some of their covariates.
For instance, smoothing breaks the relationship between a coordinate and the location error estimate around it (\texttt{VARX}, \texttt{VARY}, and \texttt{SD} in ATLAS systems, or HDOP in GPS tracking).
This makes subsequent filtering on covariates of data quality unreliable, and smoothed data are unsuitable for use with methods that model location uncertainty \citep{noonan2019, fleming2014a, fleming2020, calabrese2016}.
Furthermore, while larger $K$ may result in smoother tracks (Fig. 4.b -- e), one drawback of using a large $K$ is that short, quick forays away from the main track are likely to be smoothed away, leading to a loss in detail of the individual's small-scale movement.
Users must themselves judge how best to trade large-scale and small-scale accuracy, and choose $K$ accordingly.
One empirical way to compare $K$ is by calculating the root mean squared error (RMSE) for different $K$ on the same data.
% In general, large $K$s lead to unrealistic tracks, both because the first and last $(K - 1) / 2$ positions are handled as special cases, and because the local median approaches the global median at larger $K$.
Median smoothing is provided by the \texttt{atlastools} function \texttt{atl\_median\_smooth}, with the only option being the moving window size, which must be an odd integer (Listing 6).

@@

Thinning is implemented in \texttt{atlastools} using the \texttt{atl\_thin\_data} function, with either aggregation or resampling (specified by the \texttt{method} argument) over an interval using the \texttt{interval} argument.
The column of timestamps must be named `time' and column classes except the identity columns (\textit{see below}) should allow averaging.
The `resample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
Using `resample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
In both cases, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
Due to its development for ATLAS systems, the `aggregate' assumes the variance around coordinates is named \texttt{VARX} and \texttt{VARY}, and standard deviation around each position is named \texttt{SD}.
These columns must be present together for the function to correctly handle the error.
If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
Grouping variables' names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 7).

@@

First, users should identify positions representing bouts of stationary behaviour, for instance on their speed or residence time \citep{bracis2018}.
Patch identification assesses whether consecutive positions and the bouts they comprise are spatio-temporally independent, and clusters them together if they are not.
The splitting and lumping of positions into bouts, and bouts into patches is based on simple user specified thresholds --- the distance and the time interval between positions (and bouts) beyond which they should be considered independent.
% Patches are considered independent if either the distance or time between them exceeds their respective thresholds.
Users are encouraged to base these thresholds in the movement habits of their study species.
For example, residence patch classification of red knot movement tracks considers consecutive stationary positions independent if they are 20m apart and considers consecutive bouts independent if the distance between them $\geq$ 100m, or the time difference between them $\geq$ 30 minutes (Listing 8).
The 20m distance represents a maximum speed of 6.667 m/s between positions, above which the individual is more likely in transit.
The 100m and 30 minute thresholds are chosen to account for potentially missing data between bouts; if a track ends abruptly and then reappears $\geq$ 100m away or $\geq$ 30 minutes later, this is more safely considered a new residence patch.

>>>>

An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 3.a).
Point outliers are characterised by artificially high speeds between the outlier and the positions before and after \citep[called incoming and outgoing speed, respectively][]{bjorneraas2010}, lending a `spiky' appearance to the track.
Removing spikes is simple: remove positions with extreme incoming and outgoing speeds.
Users must first define plausible upper limits for speed and turning angle for the study species \citep{calenge2009, seidel2018}.
Here, it is important to remember that speed estimates are scale-dependent; high-throughput tracking typically overestimates the speed between positions where the animal is stationary or moving slowly due to small-scale location errors \citep{ranacher2016, noonan2019}. 
Even after data with large location errors have been removed by filters, it is advisable to begin with a liberal (high) speed threshold that excludes only the most unlikely of speeds.
Estimates of maximum speed may not always be readily obtained for all species, and an alternative is to use a data-driven threshold such as the 95\textsuperscript{th} percentile of speeds from the track.
Once a speed threshold $S$ has been chosen, positions with incoming \textit{and} outgoing speeds $\geq S$ may be identified as spikes and removed.
Some species can realistically achieve speeds $\geq S$ in fast transit segments when assisted by environmental factors, such as birds with tailwinds, and a simple filter on incoming and outgoing speeds would exclude this valid data.
To avoid removing real, fast transit segments while still excluding spikes we suggest combining the speed filter with a filter on the turning angles of each position \citep[see][]{calenge2009}.
This combined filter assumes that positions in high-throughput tracking with both high speeds and large turning angles are likely to be due to location errors, since most species are unable to turn sharply at high speed.
Users can then retain only those positions whose incoming and outgoing speeds are both $< S$ which or satisfy the condition $\theta < A$, where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
The removal of spikes is implemented using the \texttt{atl\_filter\_covariates} function (Listing 3).
Many other track metrics may be used to identify implausible movement and on which data may be filtered \citep{seidel2018}.

@@

Sometimes entire susbets of the track may be affected by the same large-scale location error.
For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 3.a, b).
These cannot be corrected by targeted removal of individual positions, as in Bjorneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
Since filtering individual positions will not suffice, algorithms to correct such errors must take a track-level view, and target the displaced sequence overall.
Track-subset algorithms are likely to be system-specific, and may be challenging to conceptualise or implement.
In the case of prolonged spikes, one relatively simple solution is identifying the bounds and removing positions between them.
% We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
Users are strongly encouraged to visualise their data before and after applying such algorithms, and we caution against relying on these method if data are heavily distorted by errors affecting entire track-subsets.

@@

Smoothing does not change the number of observations, but does decouple the coordinates from some of their covariates.
For instance, smoothing breaks the relationship between a coordinate and the location error estimate around it (\texttt{VARX}, \texttt{VARY}, and \texttt{SD} in ATLAS systems, or HDOP in GPS tracking).
This makes subsequent filtering on covariates of data quality unreliable, and smoothed data are unsuitable for use with methods that model location uncertainty \citep{noonan2019, fleming2014a, fleming2020, calabrese2016}.
Furthermore, while larger $K$ may result in smoother tracks (Fig. 4.b -- e), one drawback of using a large $K$ is that short, quick forays away from the main track are likely to be smoothed away, leading to a loss in detail of the individual's small-scale movement.
Users must themselves judge how best to trade large-scale and small-scale accuracy, and choose $K$ accordingly.
One empirical way to compare $K$ is by calculating the root mean squared error (RMSE) for different $K$ on the same data.
% In general, large $K$s lead to unrealistic tracks, both because the first and last $(K - 1) / 2$ positions are handled as special cases, and because the local median approaches the global median at larger $K$.
Median smoothing is provided by the \texttt{atlastools} function \texttt{atl\_median\_smooth}, with the only option being the moving window size, which must be an odd integer (Listing 4).

@@

Thinning is implemented in \texttt{atlastools} using the \texttt{atl\_thin\_data} function, with either aggregation or resampling (specified by the \texttt{method} argument) over an interval using the \texttt{interval} argument.
The column of timestamps must be named `time' and column classes except the identity columns (\textit{see below}) should allow averaging.
The `resample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
Using `resample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
In both cases, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
Due to its development for ATLAS systems, the `aggregate' assumes the variance around coordinates is named \texttt{VARX} and \texttt{VARY}, and standard deviation around each position is named \texttt{SD}.
These columns must be present together for the function to correctly handle the error.
If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
Grouping variables' names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 5).

@@

First, users should identify positions representing bouts of stationary behaviour, for instance on their speed or residence time \citep{bracis2018}.
Patch identification assesses whether consecutive positions and the bouts they comprise are spatio-temporally independent, and clusters them together if they are not.
The splitting and lumping of positions into bouts, and bouts into patches is based on simple user specified thresholds --- the distance and the time interval between positions (and bouts) beyond which they should be considered independent.
% Patches are considered independent if either the distance or time between them exceeds their respective thresholds.
Users are encouraged to base these thresholds in the movement habits of their study species.
For example, residence patch classification of red knot movement tracks considers consecutive stationary positions independent if they are 20m apart and considers consecutive bouts independent if the distance between them $\geq$ 100m, or the time difference between them $\geq$ 30 minutes (Listing 6).
The 20m distance represents a maximum speed of 6.667 m/s between positions, above which the individual is more likely in transit.
The 100m and 30 minute thresholds are chosen to account for potentially missing data between bouts; if a track ends abruptly and then reappears $\geq$ 100m away or $\geq$ 30 minutes later, this is more safely considered a new residence patch.
